---
title: "Working with FARS data from U.S. NHTSA"
author: "Ryan B. Harvey"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Working with FARS data from U.S. NHTSA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The U.S. National Highway Traffic Safety Administration's (NHTSA) [Fatality Analysis Reporting System](http://www.nhtsa.gov/Data/Fatality-Analysis-Reporting-System-(FARS)) provides data files detailing the factors behind traffic fatalities on United States roads. FARS is a United States-wide census providing NHTSA, Congress and the American public yearly data regarding fatal injuries suffered in motor vehicle traffic crashes.
 
The `fars` package provides tools for working with FARS data files.  This package includes FARS data files for years 2013 through 2015, but the functions will work with any CSV format data file (compressed or uncompressed) downloaded from the FARS website.

## Loading FARS data for a specific year

To load data for a specific year, say 2013, you can use the `fars_read` function with the `make_filename` function.

```{r}
library(fars)

# Load data for 2013
dat2013 <- fars_read(make_filename(2013))

# Print structure of dataset
summary(dat2013)
```

The first 10 rows look like:
```{r, echo=FALSE}
# Print first 10 rows
knitr::kable(head(dat2013, 10))
```

## Loading FARS data for multiple years

To load data for several years into a single list, you can use the `fars_read_years` function.

```{r}
# Load data for 2013-2015
years <- c(2013, 2014, 2015)
dat_all <- fars_read_years( years )

# Assign names
names(dat_all) <- years

# Inspect loaded data
str(dat_all)

# Get data frame with 2014 data
dat2014 <- dat_all[["2014"]]
```

## Summarizing FARS data for multiple years

To get a summary of the data across multiple years, you can use the `fars_summarize_data` function.

```{r}
# Summarize data for 2013-2015
summary_all <- fars_summarize_years( years )

# Inspect summary table structure
str(summary_all)

# Display summary table data
knitr::kable(summary_all)
```

## Mapping FARS data for a specific state and year

To map the FARS data for a particular state, you can use the `fars_map_state` function.  In order to use this function, you'll need to use a state code to refer to the particular state you care about.  State codes can be found in the [FARS Analytical Users Manual](https://crashstats.nhtsa.dot.gov/Api/Public/ViewPublication/812315), and are reproduced below (see "State Codes from FARS Analytical User's Manual").

```{r, fig.show="asis", fig.cap="2014 FARS data for Louisiana"}
library(maps)
library(graphics)

# Map 2014 data for Louisiana
fars_map_state(22, 2014)
```

```{r, fig.show="asis", fig.cap="2015 FARS data for Rhode Island"}
# Map 2015 data for Rhode Island
fars_map_state(44, 2015)
```

### State Codes from FARS Analytical User's Manual

State codes are reproduced here for your convenience. However, please refer to the [FARS Analytical User's Manual](https://crashstats.nhtsa.dot.gov/Api/Public/ViewPublication/812315) as the canonical source for this information.

```{r, echo=FALSE, results='asis'}
codes <- c("01","02","04","05","06","08","09","10","11","12","13","15","16","17","18","19","20","21","22","23","24","25","26","27","28","29","30","31","32","33","34","35","36","37","38","39","40","41","42","43","44","45","46","47","48","49","50","52","51","53","54","55","56")
states <- c("Alabama","Alaska","Arizona","Arkansas","California","Colorado","Connecticut","Delaware","District of Columbia","Florida","Georgia","Hawaii","Idaho","Illinois","Indiana","Iowa","Kansas","Kentucky","Louisiana","Maine","Maryland","Massachusetts","Michigan","Minnesota","Mississippi","Missouri","Montana","Nebraska","Nevada","New Hampshire","New Jersey","New Mexico","New York","North Carolina","North Dakota","Ohio","Oklahoma","Oregon","Pennsylvania","Puerto Rico","Rhode Island","South Carolina","South Dakota","Tennessee","Texas","Utah","Vermont","Virgin Islands (since 2004)","Virginia","Washington","West Virginia","Wisconsin","Wyoming")

state_codes <- data.frame(Code=codes, State=states)

knitr::kable(state_codes)
```
